---
title: "Using fredMaps to do Spatial Analysis"
author: "Kelly M. Jamrog and Jacob G. Kuruvilla"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)


```

`fredMaps` is a package designed to simplify mapping of the continental United States for beginner and intermediate R users.  This packages simplifies the graphing process from start to finish by assisting with 1) data import, 2) data management, and 3) mapping.

This document provides examples for each function within `fredMaps` using the dataframes included in the package.


## stateNames

In order to build maps with `ggplot2`, the data frame of interest needs to have a column of state names which matches the column of state names in the data frame returned by `map_data("state")`. 
The variable in `map_data("state")` is called "region", and the state names are all lowercase.

However, the problem is that the data frame we're using may not have a state names column "normalized" to the format of the "region" column in the `ggplot2` data frame. Our dataset may come with a column of state names that is abbreviated, all uppercase, capitalized, etc. Getting our column of state names to match can be a tedious and time consuming step. This is where the `stateNames` function is useful. 

The `stateName` function allows the user to specify the format of the state name and defaults to unabbreviated, lower case names to match the `ggplot2` data frame.

### Examples

Let's say we want to convert a column of uppercase state abbreviations to lowercase, unabbreviated state names. By simply passing each row to the function `stateNames`, we can convert it to the form we want.

```{r, message=FALSE, warning=FALSE}
library(fredMaps)
# on a single observation
stateNames("NY")

# on a vector containing state names
stateNames(c("New york", "Boston", "ct"), tolower = FALSE, abb = TRUE, replace.na = TRUE)
```


Note: if a non-state is inputted into `stateNames`, the user can specificy if they want it replaced replace with NAs or have it return without any changes made. 



## mergeOnState

An important step when working with data is merging two data frames.  This step can become tedious if the data is obtained from different sources and the column to merge on is not reported in the same format. This can often be the case when working with state data since states can be referenced by their full name or abbrevation, and with different types of capitalization. 

`mergeOnState` allows users to merge two datasets on a column that contains state names or abbreviations and does not require that the columns are in the same format. 


Let's see how this function works using `gdp` and `bingedrinking`, two data frames that come with `fredMaps`.

First, let's look at the data frames.

```{r}
library(fredMaps)

head(gdp)

head(bingedrinking)

```

`gdp` is a dataframe with two columns:

* X, which contains state names and

* gdpPerCap, which contains nominal GDP per capita in a given state for the year 2015.

`bingedrinking` is a dataframe with three columns:

* LocationAbbr, which contains state names as abbrevations, 

* BingeDrinking, which contains binge drinking prevalence among adults aged >= 18 years, and

* Group, which indicates a subset of the population (Overall, Male, Female)


Normally, prior to merging, it would be necessary to convert either the abbrevations to state names or vice versa and to make sure that the two columns are in the same case. However, `mergeOnState` allows the user to do this all in one step.  

We are ready to merge!

```{r}
drinking_GDP <- mergeOnState(bingedrinking, LocationAbbr, gdp, X, merge = "inner")
head(drinking_GDP)
```

Now that we have our data merged, we can move on to our analysis! Woo!

```{r}
cor(drinking_GDP$BingeDrinking, drinking_GDP$gdpPerCap)
```


## `fredMaps` and `fredMapsPoint`

Now that we are done with our data management, let's do some plotting.

`fredMaps` and `fredMapsPoint` are functions designed to make plotting quick and easy.  Regular `ggplot2` code for spatial analysis can be lengthy and complex even when creating a basic plot.  These functions are designed to enable the user to map their state-level data in one step. 


### `fredMaps`

`fredMaps` works by taking 6 arguments which are supplied to what would have otherwise been cumbersome `ggplot2` code: a data frame, a variable in the dataframe with values to plot, a color representing lower values, a color representing higher values, and a list of states. 

Using `ggplot2` functions `geom_map` and `map_data`, which extracts shape files for the continental United States, `fredMaps` prints the spatial polygons representing states, which are filled with user specificed colors. 

Unlike `fredMaps`, `fredMapsPoint` plots the map of the continental United States with points whose size represents a numeric variable in a data frame and color represents another numeric, factor, or character variable in the data frame. In addition, the function also takes the variable in the data frame containing latitudes and longitudes for points.


Let's work through some examples together.

##Example: Plotting Adult Obesity Rates


Let's use the `obesity` data set to plot average adult obesity rates by state.  States with lower values of adult obesity rate are colored pink, and states with higher values of adult obesity are colored red. Feel free to change the colors of the plot by change the arugements `low` and `high`.  The `fredMaps` function will allow us to do this in one line:

```{r, message=FALSE, warning=FALSE, fig.align= 'center', dpi = 72, fig.width= 9, fig.height= 7}
fredMaps(obese, AdultsObese, State, low="pink", high="red")
```

As seen from our plot, Mississippi has the highest average adult obesity rate.


In some cases we may only want a subset of states. For example, let's say we want to compare average child obesity rates only for states in the midwest. We can do this by specifying the `states` argument in `fredMaps`: 

```{r, message=FALSE, warning=FALSE, fig.align= 'center', dpi = 72, fig.width= 9, fig.height= 7}
fredMaps(obese, Children, State, low = "pink", high = "purple",
         states = c("illinois", "indiana", "iowa", "kentucky",
                                           "michigan", "minnesota","missouri",
                                           "north dakota", "ohio", "south dakota",
                                           "wisconsin"))
```

Zoning on midwestern states, we can see that Minnesota has the lowest average childhood obesity rate.

While visualizing data by filling states with all kinds of colors can be helpful, a potential issue with these kinds of maps is that the human eye tends to associate larger shapes with with larger magnitudes or importance. Hence, one may associate larger states with higher obesity rates. To get around this problem, we can overlay points on the map instead of coloring the states.  The size and color of these points can help us capture additional information.


##Example: Plotting School Enrollment Rates

Let's use the `school_enrollments` dataset to visualize the total number of students enrolled as well as school labels by state. By inputting these variables and the variable names for latitudes and longitudes into `fredMapsPoint`, we can do just that:

```{r, message=FALSE, warning=FALSE, fig.align= 'center', dpi = 72, fig.width= 9, fig.height= 7}
fredMapsPoint(college_enrollment, size=enrollment, color=label, long=long, lat=lat)
```



# `fredMaps` for Intermediate Users

## User Provided Shapefiles

`fredMaps` was built to help users map the continental United States.  However, often, one might be interested in plotting non-standard geographical locations.  If a user has a shapefile, they can use `shapeToDF`, `shapeToMap`, and `dfToMap` to plot their own data.

For reference, a shapefile is a group of files containing feature geometry and feature attribute data.  There are three main feature geometry types: points, lines, and polygons.  `fredMaps` allows users to plot shapefiles containing polygons.  

There are many free, online sources for shapefiles.  Below are some websites that have free shapefiles available to download:

* [Neighborhoods in New York City](https://www1.nyc.gov/site/planning/data-maps/open-data/dwn-nynta.page)

* [Historical urban ecological data for cities in the United States](http://uadata.org/hue/)

* [Estuaries, Lakes, and Ecoregions in Utah](https://archive.epa.gov/nheerl/arm/web/html/design_files.html)

When downloaded, shapefiles are stored as a zip file that must be extracted.  At minimum, the folder must contain a .shp file, however it may often contain additional files.  

## `shapeToDF`, `shapeToMap`, and `dfToMap`

`shapeToDF`, `shapeToMap`, and `dfToMap` simplify the process of reading in shapefiles and mapping using ggplot.  Both `shapeToDF` and `shapeToMap` require the user to input the path to the folder containing the shapefile and the name of the shapefile **without** an extention.

Let's use these functions to plot New York City neighborhoods using a shapefile that can be downloaded from the first link above.

The shapefile that we will be using is stored in a folder called `nynta_18d`.  If we look inside the folder, there are 5 files:

* nynta.shp.xml

* nynta.prj

* nynta.shx

* nynta.dbf

* nynta.shp

Notice that all of the files contain the same "name", but have different file extensions.  We will pass this "name" into the `shapeToMap` and `shapeToDF` function as the `file` argument.

### `shapeToMap`

For a quick graph, let's use `shapeToMap` to read in the shapefiles and create a graph all in one step. 


```{r fig.align= 'center', dpi = 72, fig.width= 7, fig.height= 7}
shapeToMap("nynta_18d", "nynta")
```

Tah dah! A map of New York City!

### `shapeToDF` and `dfToMap`

For users interested in a little more control, `shapeToDF` and `dfToMap` are a great option.  

`shapeToDF` reads in shapefiles and outputs a dataframe that can be fed to `dfToMap`. One of the benefits of doing this process in two steps is that the user can specify a column to be used to color the map.  

If the user prefers even more control, they can use the output of `shapeToDF` with their own `ggplot2` code.

Finally, both `shapeToMap` and `dfToMap` return a `ggplot2` object which allows users to overlay additional ggplot functions, including changing axis labels, titles, and colors.


## `shapeToDF` and `dfToMap`

Let's again plot New York City neighborhoods, but this time use `shapeToDF` and `dfToMap` to color in the different boroughs.

```{r}
ny <- shapeToDF("nynta_18d", "nynta")
names(ny)
```


By looking at our dataframe, we notice that we have a BoroName column which we can use to fill in our map. Let's also personalize this map by adding a title and label for the legend.


```{r fig.align= 'center', dpi = 72, fig.width= 7, fig.height= 7}
library(ggplot2)

dfToMap(ny, ny$BoroName) +
  labs(title = "New York City", fill = "Borough") 
```


